# %% [code]
---
title: "Global Suicide Trends & Analysis"
author: 'John Okoye'
date: '02 Apr 2023'
output:
  html_document:
    number_sections: true
    toc: true
    code_folding: "hide"
    theme: readable
    highlight: haddock
---
  
  
# **Introduction: About Dataset**
### Content  
This compiled dataset pulled from four other datasets linked by time and place, and was built to find signals correlated to increased suicide rates among different cohorts globally, across the socio-economic spectrum.

### References  
United Nations Development Program. (2018). Human development index (HDI). Retrieved from http://hdr.undp.org/en/indicators/137506
World Bank. (2018). World development indicators: GDP (current US$) by country:1985 to 2016. Retrieved from http://databank.worldbank.org/data/source/world-development-indicators#
[Szamil]. (2017). Suicide in the Twenty-First Century [dataset]. Retrieved from https://www.kaggle.com/szamil/suicide-in-the-twenty-first-century/notebook
World Health Organization. (2018). Suicide prevention. Retrieved from http://www.who.int/mental_health/suicide-prevention/en/
  
### Inspiration  
Suicide Prevention.

  

## Import and Clean  
***Data Cleaning Change Log***  
  
* Changed column names for ease of reference (gdp_per_capita, gdp_for_year).  
* 2016 data was removed due to missing data. Few countries had any and those that did often had data missing.  
* HDI was removed due to missing data for some years and countries.  
* Removed country-year column due to duplication.  
* Removed suicides/ 100k. A more accurate calculation will be done.  
* Generation variable is not consistent.  
* Continent was added to the dataset using the `countrycode` package.  


```{r message = FALSE, warning = FALSE}
library(tidyverse) #General  
library(dplyr) #Data Manipulation  
library(janitor) #Data cleaning package  
library(countrycode) #To create column for continents  
library(ggthemes) #For plot themes  
library(gridExtra) #For grid arranged plots  
library(rworldmap) #For map

#Import Data
data_set <- read.csv("/kaggle/input/suicide-rates-overview-1985-to-2016/master.csv")

#Selecting columns and cleaning columns
data <- data_set %>% 
select(-"HDI.for.year", -"suicides.100k.pop", -"country.year") %>% 
rename(gdp_for_year = 7, gdp_per_capita = 8)

data$age <- gsub(" years", "", data$age)
data$sex <- ifelse(data$sex == "male", "Male", "Female")

#Creating column for continents
data$continent <- countrycode(sourcevar = data$country, origin = "country.name", destination = "continent")

#Making age variable ordinal
data$age <- factor(data$age, 
                   ordered = T, 
                   levels = c("5-14",
                              "15-24", 
                              "25-34", 
                              "35-54", 
                              "55-74", 
                              "75+"))

#Making generation variable ordinal
data$generation <- factor(data$generation, 
                          ordered = T, 
                          levels = c("G.I. Generation", 
                                     "Silent",
                                     "Boomers", 
                                     "Generation X", 
                                     "Millenials", 
                                     "Generation Z"))

#Global average as a yardstick
global_average <- (sum(as.numeric(data$suicides_no)) / sum(as.numeric(data$population))) * 100000

#Default theme
theme <- theme(text = element_text(family = "serif", size = 12),legend.position = "none")
```

# **Global Analysis**
```{r message = FALSE, warning = FALSE}
data %>%
  group_by(year) %>%
  summarize(population = sum(population), 
            suicides = sum(suicides_no), 
            suicides_per_100k = (suicides / population) * 100000) %>%
  ggplot(aes(x = year, y = suicides_per_100k)) + 
  geom_line(col = "#12a4d9", size = 1) + 
  geom_point(col = "#12a4d9", size = 2) + 
  geom_hline(yintercept = global_average, linetype = 2, color = "grey40", size = 1) +
  labs(title = "Global Suicides (per 100k)",
       subtitle = "Trend over time, 1985 - 2015.",
       caption = "Data Source: World Health Organization",
       x = "Year", 
       y = "Suicides per 100k") + 
  scale_x_continuous(breaks = seq(1985, 2015, 2)) + 
  scale_y_continuous(breaks = seq(10, 20)) + theme +
  annotate("text", x = 2003, y = 12.5, label = "Global Average", vjust = 1, size = 3, color = "grey40") +
  annotate(
    "curve",
    x = 2003, y = 12.5,
    xend = 2003, yend = 13.1,
    arrow = arrow(length = unit(0.2, "cm"), type = "closed"),
    color = "grey40"
  )
```

***Insights***  

* Upward trend between **1991** to **1995**  
* Peak suicide rate was **15.3** deaths per 100k in **1995**  
* Decreased steadily, to **11.5** per 100k in **2015** (**~25% decrease**)  


# **By Continent**
```{r message = FALSE, warning = FALSE}
#Grouping by continent and arranging so the plot is orderly 
continent <- data %>%
  group_by(continent) %>%
  summarize(suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000) %>%
  arrange(suicide_per_100k)

continent$continent <- factor(continent$continent, ordered = T, levels = continent$continent)

continent_plot <- ggplot(continent, aes(x = continent, y = suicide_per_100k, fill = continent)) + 
  geom_col() + 
  labs(title = "Global Suicides (per 100k) by Continent",
       x = "Continent", 
       y = "Suicides per 100k", 
       fill = "Continent") + 
  scale_y_continuous(breaks = seq(0, 20, 1), minor_breaks = F) + theme

#With Year
continent_time <- data %>%
  group_by(year, continent) %>%
  summarize(suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000)

#To set an order to categorical variable
continent_time$continent <- factor(continent_time$continent, ordered = T, levels = continent$continent)

continent_time_plot <- ggplot(continent_time, aes(x = year, y = suicide_per_100k, col = factor(continent))) + 
  facet_grid(continent~., scales = "free_y") + 
  geom_line() + 
  geom_point() + 
  labs(title = "Trends Over Time (per 100k) by Continent", 
       x = "Year", 
       y = "Suicides per 100k", 
       color = "Continent") +
  scale_x_continuous(breaks = seq(1985, 2015, 5), minor_breaks = F) + theme

grid.arrange(continent_plot, continent_time_plot, ncol = 2)
```

***Insights***  

* European rate highest overall, but has steadily **decreased by 40% since 1995**  
* The **European rate for 2015 similar to Asia & Oceania**  
* The trendline for Africa is due to poor data quality - just 3 countries have provided data  
* **Oceania & Americas are trending upwards which is concerning**  
  
  
# **By Sex**
```{r message = FALSE, warning = FALSE}
sex_plot <- data %>%
  group_by(sex) %>%
  summarize(suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000) %>%
  ggplot(aes(x = sex, y = suicide_per_100k, fill = sex)) + 
  geom_col() + 
  labs(title = "Global suicides (per 100k) by Sex",
       x = "Sex", 
       y = "Suicides per 100k") +
  scale_y_continuous(breaks = seq(0, 25), minor_breaks = F) + 
  theme

#With Year
sex_time_plot <- data %>%
  group_by(year, sex) %>%
  summarize(suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000) %>%
  ggplot(aes(x = year, y = suicide_per_100k, col = sex)) + 
  facet_grid(sex ~ ., scales = "free_y") + 
  geom_line() + 
  geom_point() + 
  labs(title = "Trends Over Time, by Sex", 
       x = "Year", 
       y = "Suicides per 100k", 
       color = "Sex") +
  scale_x_continuous(breaks = seq(1985, 2015, 5), minor_breaks = F) + theme

grid.arrange(sex_plot, sex_time_plot, ncol = 2)
```

***Insights***  

* Globally, the rate of suicide for men has been **~3.5x higher for men**  
* Both **male & female suicide rates peaked in 1995** and has been declining since  
* This ratio of 3.5 : 1 (male : female) has remained relatively constant since the mid 90s  


```{r message = FALSE, warning = FALSE}
#Sex by continent
data %>%
  group_by(continent, sex) %>%
  summarize(n = n(), 
            suicides = sum(as.numeric(suicides_no)), 
            population = sum(as.numeric(population)), 
            suicide_per_100k = (suicides / population) * 100000) %>%
  ggplot(aes(x = continent, y = suicide_per_100k, fill = sex)) + 
  geom_bar(stat = "identity", position = "dodge") + 
  geom_hline(yintercept = global_average, linetype = 2, color = "grey40", size = 1) +
  theme + theme(legend.position = "bottom", legend.direction = "horizontal") +
  annotate("text", x = 1, y = 20, label = "Global Average", vjust = 1, size = 3, color = "grey40") +
  annotate(
    "curve",
    x = 1, y = 20,
    xend = 1.5, yend = 13.5,
    arrow = arrow(length = unit(0.2, "cm"), type = "closed"),
    color = "grey40") +
  labs(title = "Gender Disparity by Continent",
       x = "Continent", 
       y = "Suicides per 100k", 
       fill = "") +
  coord_flip()
```

***Insights***   

* European men were at the highest between 1985 - 2015 at almost 30 suicides per 100k  
* The Americas had the lowest cases of men suicides(excluding Africa due to missing data countries) with a litle over the global avergae of 13.1 per 100k  
* Europe’s rate was at 2.3x higher than global average   

```{r message = FALSE, warning = FALSE}
#Age
age_plot <- data %>%
  group_by(age) %>%
  summarize(suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000) %>%
  ggplot(aes(x = age, y = suicide_per_100k, fill = age)) + 
  geom_col() + 
  labs(title = "Global suicides per 100k by Age",
       x = "Age", 
       y = "Suicides per 100k") +
  scale_y_continuous(breaks = seq(0, 30, 1), minor_breaks = F) + theme

#With Year
age_time_plot <- data %>%
  group_by(year, age) %>%
  summarize(suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000) %>%
  ggplot(aes(x = year, y = suicide_per_100k, col = age)) + 
  facet_grid(age~. , scales = "free_y") + 
  geom_line() + 
  geom_point() + 
  labs(title = "Trends Over Time (per 100k) by Age", 
       x = "Year", 
       y = "Suicides per 100k", 
       color = "Age") + 
  scale_x_continuous(breaks = seq(1985, 2015, 5), minor_breaks = F) + theme

grid.arrange(age_plot, age_time_plot, ncol = 2)
```

***Insights***  

* Globally, suicide increases with age  
* Since 1995, suicide rate for everyone aged >= 15 has been linearly decreasing  
* The suicide rate of those aged 75+ has dropped by more than 50% since 1990  
* Suicide rate in the ‘5-14’ category remains roughly static and small (< 1 per 100k per year)  

```{r message = FALSE, warning = FALSE}
#Age group by continent
data %>%
  group_by(continent, age) %>%
  summarize(suicides = sum(as.numeric(suicides_no)), 
            population = sum(as.numeric(population)), 
            suicide_per_100k = (suicides / population) * 100000) %>%
  ggplot(aes(x = continent, y = suicide_per_100k, fill = age)) + 
  geom_bar(stat = "identity", position = "dodge") + 
  geom_hline(yintercept = global_average, linetype = 2, color = "grey40", size = 1) +
  labs(title = "Age Disparity by Continent",
       x = "Continent", 
       y = "Suicides per 100k", 
       fill = "Age") + 
  annotate("text", x = 1.9, y = 23.5, label = "Global Average", vjust = 1, size = 3, color = "grey40") +
  annotate(
    "curve",
    x = 1.5, y = 23.1,
    xend = 1.5, yend = 13.1,
    arrow = arrow(length = unit(0.2, "cm"), type = "closed"),
    color = "grey40") + 
  theme(text = element_text(family = "serif", size = 12))
```

***Insights***  

* For the Americas, Asia & Europe (which make up most of the dataset), suicide rate increases with age  
* Oceania & Africa’s rates are highest for those aged 25 - 34  

# **By Country**
```{r fig.width = 8, fig.height = 15}
country <- data %>%
  group_by(country, continent) %>%
  summarize(suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000) %>%
  arrange(desc(suicide_per_100k))

country$country <- factor(country$country, 
                          ordered = T, 
                          levels = rev(country$country))


ggplot(country, aes(x = country, y = suicide_per_100k, fill = continent)) + 
  geom_col() + 
  geom_hline(yintercept = global_average, linetype = 2, color = "grey40", size = 1) +
  coord_flip() +
  labs(title = "Global suicides per 100k, by Country",
       x = "Country", 
       y = "Suicides per 100k", 
       fill = "Continent") +
  scale_y_continuous(breaks = seq(0, 45, 2),expand = c(0,0)) + 
  theme + theme(legend.position = "bottom", legend.direction = "horizontal") +
  labs(fill = "") + 
  annotate("text", x = 50, y = 20, label = "Global Average", vjust = 1, size = 3, color = "grey40") +
  annotate(
    "curve",
    x = 50, y = 20,
    xend = 50, yend = 13.1,
    arrow = arrow(length = unit(0.2, "cm"), type = "closed"),
    color = "grey40")
```

***Insights***  

* Lithuania’s rate has been highest by a large margin: > 41 suicides per 100k, per year  
* Large over representation of European countries with high rates, few with low rates  

Below is a geographical heat map of the suicide rates between the timeframe of this analysis - note the lack of data for Africa and Asia, and bear in mind that 7 countries have been removed due to insufficient data.

```{r message = FALSE, warning = FALSE}
#Country heat map
country <- data %>%
  group_by(country) %>%
  summarize(suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000)

countrydata <- joinCountryData2Map(country, joinCode = "NAME", nameJoinColumn = "country")
```

```{r message = FALSE, warning = FALSE}
par(mar=c(0, 0, 0, 0)) # margins
mapCountryData(countrydata, 
               nameColumnToPlot="suicide_per_100k", 
               mapTitle="", 
               colourPalette = "heat", 
               oceanCol="#5accd0", 
               missingCountryCol="grey", 
               catMethod = "pretty")
```

```{r message = FALSE, warning = FALSE}
country_mean_gdp <- data %>%
  group_by(country, continent) %>%
  summarize(suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000, 
            gdp_per_capita = mean(gdp_per_capita))

ggplot(country_mean_gdp, aes(x = gdp_per_capita, y = suicide_per_100k, col = continent)) + 
  geom_point() + 
  scale_x_continuous(labels=scales::dollar_format(prefix="$"), breaks = seq(0, 70000, 10000)) + 
  labs(title = "Correlation between GDP (per capita) and Suicides per 100k",
       x = "GDP (per capita)", 
       y = "Suicides per 100k", 
       col = "Continent") +
  theme(text = element_text(family = "serif", size = 12))
```

# **Key Insights**

* Suicide rates are decreasing globally after a spike in 1995.  
* On average, suicide rate increases with age.  
* There is a weak positive relationship between a countries GDP (per capita) and suicide rate.  
* There is an overrepresentation of men in suicide deaths at every level of analysis (globally, at a continent and country level). Globally, the male rate is at 3.5x higher.  


